Analysis-ready VCF at Biobank-scale using Zarr

### 2024-05-07 paper.tex
Nother intro pass

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, storing information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.

VCF's simple text-based design and widespread support makes it an 
excellent archival format, but it is a poor basis for computational
workflows. 
% Methods that require efficient access to genotype data
% either require convertion to the
% PLINK~\cite{purcell2007plink,chang2015second} 
% or BGEN~\citep{band2018bgen} 
% formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
% or use 
% bespoke binary formats~\citep[e.g.][]{
% % Uses custom "bref3" format,
% % https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
% browning2018one, 
% % .samples Zarr format
% kelleher2019inferring,
% % Has a "xcftools" package, but it still looks pretty experimental
% hofmeister2023accurate} 
% that support the required access patterns.
% While PLINK and BGEN support efficient access to genotype data, neither
% can accommodate the full flexibility of the VCF data model and conversion
% is lossy.
Processing of Biobank-scale datasets can be split into a 
few broad categories. The first level of analysis that is 
done is quality control (QC). Variant calling QC is an 
involved and multi-faceted 
task [CITATIONS?], requiring analysis of many different metrics
and incurring substantial computation.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets.
Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
Hail distributes data and calculations over a cluster, enabling
interactive analysis of very large datasets. It has been successfully used 
in many studies, and instrumental in datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
After QC, a common task is to perform a genome wide association study
(GWAS)~\cite{uffelmann2021genome}. The majority of tools for 
performing GWAS and associated tasks require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional files.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF. 
For example, at two bits 
per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'
Many applications examine data on a variant-by-variant
basis, but others,
e.g.\ reference-panel based phasing and imputation 
methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
% and Ancestral Recombination Graph 
% (ARG) inference~\cite{brandt2024promise,lewanski2024era,wong2024general}
require access on a sample-by-sample basis.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or samples of interest for a particular 
study (see section XXX).

VCF (and BCF) cannot support any of these workflows efficiently
at the Biobank scale, and workarounds of creating subsets 
of data (i.e., discarding QC metrics in VCF) or copying to 
PLINK format are becoming less effective.
The most intrinsically limiting aspect of VCF's design is its 
row-wise layout of data within a file, which means that (for example)
data for a particular sample cannot be efficiently retrieved.
The file-oriented paradigm is also unsuited to the realities 
of modern datasets, which are too large to download and 
often required to stay in-situ by data-access agreements.
Large files are stored in cloud environments, where the 
file systems that are required by classical file-oriented tools
are expensively emulated on the basic building blocks
of object storage. 
These multiple layers of inefficiences around processing
VCF data at scale in the cloud mean that it is 
time-consuming and expensive, effectively making 
% Dunno if we want to get into this, but it's true
analysis of these datasets the preserve of a well-funded elite.

The analysis of Biobank-scale variation 
datasets requires a primary data representation that can 
efficiently support different types of query without 
conversion or intermediate copies, and operate 
directly from cloud object stores. Such a representation
can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For such a representation to be FAIR~\citep{wilkinson2016fair},
it must also \emph{accessible}: ``open, free, and universally
implementable''.
There is no currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
While Hail is powerful software its file formats are not documented or 
designed for reuse.
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

Here, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be 
efficiently stored and analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large multidimensional data as a regular
grid of compressed chunks. 
Zarr's simplicity and its direct mapping of chunks to 
keys in an object store have lead to 
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification which formalises this 
mapping, and the \texttt{vcf2zarr} 
utility which reliably converts large-scale VCFs into Zarr 
format. We show that VCF Zarr is much more compact than 
VCF (and BCF), and is competitive with state-of-the-art
VCF compression tools. 
Moreover, we show that Zarr's storage of data in an analysis-ready 
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competetive
with state-of-the-art file-oriented methods. Finally, we show the 
utility of VCF Zarr on the Genomics England aggV2 dataset,
demonstrating that common \texttt{bcftools} queries can be performed orders
of magnitude more quickly using simple Python scripts.

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, encoding information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale initiatives such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.

VCF's simple text-based design and widespread support makes it an 
excellent archival format, but it is a poor basis for computational
workflows. 
Methods that require efficient access to genotype data
either require convertion to the
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
formats~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
bespoke binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy.
PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format provides efficient access to genotype 
data, but file sizes are substantially larger than 
the equivalent compressed VCF. 
For example, at two bits 
per genotype, the full genotype matrix for the GraphTyper SNP dataset 
in the UKB WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

Processing of Biobank-scale datasets can be split into a 
few broad categories. The most basic analysis 
is quality control (QC). Variant calling QC is an 
involved and multi-faceted 
task [CITATIONS?], requiring analysis of many different metrics
and incurring substantial computation.
Hail~\cite{ganna2016ultra,hail2024} has become the dominant platform
for quality control of large-scale variation datasets.
Building on components from the Hadoop ecosytem~\citep{white2012hadoop},
Hail distributes data and calculations over a cluster, enabling
interactive analysis of very large datasets. It has been successfully used 
in many studies, and instrumental in datasets such as 
gnomadAD~\cite{karczewski2020mutational,chen2024genomic}.
A common task to perform after QC is 
a genome wide association study
(GWAS)~\cite{uffelmann2021genome}. The majority of tools for 
performing GWAS and related analyses require
data to be in PLINK format~\cite[e.g][]{chang2015second} [MORE],
and so data must be ``hard-called'' according to some QC criteria
and exported to additional files.
Genotype calls are often further refined via statistical methods,
% methods~\citep[e.g.][MORE and CHECK]{browning2021fast,browning2023statistical}
for example by phasing~[CITES] and imputation[CITES], usually 
creating an additional copy of the dataset.
Finally, variation datasets are often queried in exploratory
analyses, to find regions or samples of interest for a particular 
study (see section XXX for an example).

VCF cannot support any of these workflows efficiently
at the Biobank scale, and workarounds of copying subsets 
of data (i.e., discarding QC metrics in VCF) or copying to 
PLINK format have reached their limits.
The most intrinsically limiting aspect of VCF's design 
is its row-wise layout of data, which means that (for example)
information for a particular sample or field cannot be obtained without
retrieving the entire dataset.
The file-oriented paradigm is also unsuited to the realities 
of modern datasets, which are too large to download and 
often required to stay in-situ by data-access agreements.
Large files are stored in cloud environments, where the 
file systems that are required by classical file-oriented tools
are expensively emulated on the basic building blocks
of object storage. 
These multiple layers of inefficiences around processing
VCF data at scale in the cloud mean that it is 
time-consuming and expensive, 
% Can easily delete this last bit:
effectively making 
analysis of these datasets the preserve of a well-funded elite.

To fully utilise these vast variation datasets,
we require a primary data representation that can 
support different types of query without 
conversion or intermediate copies, and operate 
directly from cloud object stores. Such a representation
can be termed ``analysis-ready''
and ``cloud-native''~\citep{abernathey2021cloud}.
For the representation to be FAIR~\citep{wilkinson2016fair},
it must also \emph{accessible} using protocols that are 
``open, free, and universally implementable''.
There is no currently no efficient, FAIR representation of genetic variation
data suitable for cloud deployments.
While Hail is powerful software its file formats are not documented or 
designed for reuse.
Similarly, commerical solutions that have emerged to facilitate
the analysis of large-scale genetic variation data are either
based on proprietary~[Terra? BigQuery?]
or single-vendor technologies~\cite[e.g.][]{tiledb2024}.
The next generation of VCF analysis methods requires
an open, free and transparent data representation 
with multiple independent implementations.

In this article, we decouple the VCF data model from its row-oriented
file definition, and show how the data can be 
efficiently stored and analysed in a cloud-native, FAIR manner.
We do this by translating VCF data into Zarr format,
a method of storing large multidimensional data as a regular
grid of compressed chunks. 
Zarr's elegant simplicity and direct mapping of chunks to 
keys in an object store have lead to 
it gaining substantial traction
across the sciences, and it is now used in multiple petabyte-scale
datasets in cloud deployments (see Methods for details).
We present the VCF Zarr specification which formalises this 
mapping, and the \texttt{vcf2zarr} 
utility which reliably converts large-scale VCFs into Zarr 
format. We show that VCF Zarr is much more compact than 
VCF and is competitive with state-of-the-art
VCF compression tools. 
Moreover, we show that Zarr's storage of data in an analysis-ready 
format greatly facilitates computation,
with various benchmarks being substantially faster than
\texttt{bcftools} based pipelines, and again competetive
with state-of-the-art file-oriented methods. Finally, we show the 
utility of VCF Zarr on the Genomics England aggV2 dataset,
demonstrating that common \texttt{bcftools} queries can be performed orders
of magnitude more quickly using simple Python scripts.
